---
title: "cellNexus"
author: "Mangiola et al."
package: cellNexus
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
params:
  demo_metadata: true
vignette: >
  %\VignetteIndexEntry{cellNexus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_format = BiocStyle::html_document(toc = TRUE, toc_float = TRUE),
            params = list(
                demo_metadata = TRUE
            )
        )
        rmarkdown::render(
            x,
            output_file = "README.md",
            output_format = "github_document",
            output_dir = proj_root,
            knit_root_dir = proj_root,
            params = list(
                demo_metadata = TRUE
            )
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
root_dir <- knitr::opts_knit$get("root.dir")
if (!is.null(root_dir)) {
  # This hack fixes the relative image paths.
  # See https://github.com/rstudio/rmarkdown/issues/2473
  knitr::opts_knit$set(
    output.dir = root_dir
  )
}
proj_root <- rprojroot::find_package_root_file() |>
  normalizePath()
# Utility function for figures to force them to have the correct path
find_figure <- function(names) {
  rprojroot::find_package_root_file() |>
    file.path("man", "figures", names)
}
METADATA_URL <- if (params$demo_metadata) {
  cellNexus::SAMPLE_DATABASE_URL
} else {
  cellNexus::get_metadata_url
}
```

<!-- badges: start -->
[![Lifecycle:maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#maturing)
<!-- badges: end -->

# Introduction

`cellNexus` builds upon and extends the functionality of the previously released `CuratedAtlasQueryR`, providing a unified query and access interface to the harmonised, curated, and reannotated CELLxGENE human cell atlas. It enables reproducible and programmatic exploration of large-scale single-cell data resources, supporting retrieval at the cell, sample, and dataset levels through flexible filtering by tissue, cell type, experimental condition, or other metadata features. The retrieved data are returned for downstream analysis.

`cellNexus` integrates over 40 million human cells processed with standardised quality control, consistent normalisation, and unified abundance representations—including single-cell, counts-per-million, normalised expression, and pseudobulk layers. This harmonised design facilitates efficient cross-dataset analyses and downstream integration.

Data are hosted on the ARDC Nectar Research Cloud, and most `cellNexus` functions interact with Nectar via web requests, so a 
network connection is required for most functionality.  

`cellNexus` builds on top of package CuratedAtlasQueryR. While both rely on 
pre-computed expression layers, they differ in how these layers are generated.
cellNexus implements a more standardised workflow, including explicit 
removal of empty droplets and dead cells, followed by harmonised quality control, 
normalisation, and multi-layer data generation. Through this process, it 
produces updated datasets that remain aligned with the evolving CELLxGENE releases.

```{r fig-logo, echo=FALSE, out.height = c("139px"), out.width = "120x" }
find_figure("logo.png") |>
  knitr::include_graphics()
```

```{r fig-funders, echo=FALSE, out.height = c("58px"), out.width = c("155x", "129px", "202px", "219px", "180px")}
c(
  "svcf_logo.jpeg",
  "czi_logo.png",
  "bioconductor_logo.jpg",
  "vca_logo.png",
  "nectar_logo.png"
) |>
  find_figure() |>
  knitr::include_graphics()
```

# Query interface

## Installation

```{r install-dev, eval=FALSE}
devtools::install_github("MangiolaLaboratory/cellNexus")
```

## Load the package

```{r load-cellnexus, message=FALSE, warning=FALSE}
library(cellNexus)
```

## Load additional packages

```{r load-ggplot2}
suppressPackageStartupMessages({
  library(ggplot2)
})
```

## Load and explore the metadata

### Load the metadata
By default, `get_metadata()` loads harmonised annotations. Users can retrieve original Census annotations by the function `join_census_table()`. 

```{r load-metadata-demo, eval=FALSE}
metadata <- get_metadata() |>
  join_census_table()
metadata
```

```{r load-metadata-check, echo=FALSE}
# Note: a custom cache is used here ONLY for R CHECK compliance purposes. Users will NOT need to specify a custom cache
metadata <- get_metadata(
  cloud_metadata = METADATA_URL["cellnexus"],
  cache_directory = tempdir()
) |>
  join_census_table(cloud_metadata = METADATA_URL["census"])
metadata
```

Metadata is saved to `get_default_cache_dir()` unless a custom path is provided via the cache_directory argument. The `metadata` variable can then be re-used for all subsequent queries. 

### Explore the tissue 

```{r explore-tissue}
metadata |>
  dplyr::distinct(tissue, cell_type_unified_ensemble)
```

## Quality control
cellNexus metadata applies standardised quality control to filter out empty droplets, dead or damaged cells, doublets, and samples with low gene counts.

```{r qc-filter}
metadata <- metadata |>
  keep_quality_cells()

metadata <- metadata |>
  dplyr::filter(feature_count >= 5000)
```


## Download single-cell RNA sequencing counts 

### Query raw counts

```{r get-sce-counts}
single_cell_counts <-
  metadata |>
  dplyr::filter(
    self_reported_ethnicity == "African" &
      assay == "10x 3' v3" &
      tissue == "lung parenchyma" &
      cell_type_unified_ensemble == "cd14 mono"
  ) |>
  get_single_cell_experiment()

single_cell_counts
```

### Query counts scaled per million

```{r get-sce-cpm}
single_cell_cpm <-
  metadata |>
  dplyr::filter(
    self_reported_ethnicity == "African" &
      assay == "10x 3' v3" &
      tissue == "lung parenchyma" &
      cell_type_unified_ensemble == "cd14 mono"
  ) |>
  get_single_cell_experiment(assays = "cpm")

single_cell_cpm
```

### Query SCT normalised counts
```{r get-sce-sct}
single_cell_sct <-
  metadata |>
  dplyr::filter(
    self_reported_ethnicity == "African" &
      assay == "10x 3' v3" &
      tissue == "lung parenchyma" &
      cell_type_unified_ensemble == "cd14 mono"
  ) |>
  get_single_cell_experiment(assays = "sct")

single_cell_sct
```

### Query pseudobulk

```{r get-pseudobulk}
pseudobulk_counts <-
  metadata |>
  dplyr::filter(
    assay == "10x 5' v1" &
      tissue == "lung" &
      cell_type_unified_ensemble == "treg"
  ) |>
  get_pseudobulk()

pseudobulk_counts
```

## Download cell communication metadata

Cell communication metadata was generated based on post-QC cells per sample using `CellChat v2` method. It uses our harmonised cell type annotation (cell_type_unified_ensemble) to infer the communication. It captures inferred communication at both the ligand–receptor pair level and the signalling pathway level.  

interaction_count: The number of inferred interactions between each pair of cell groups.  
interaction_weight: The aggregated communication strength between each pair of cell groups.  

For definitions of additional annotations, please refer to the CellChat v2 documentation: https://github.com/jinworks/CellChat.

```{r get-cell-communication, eval=FALSE}
get_cell_communication_strength()
```

```{r get-cell-communication-demo, echo=FALSE}
get_cell_communication_strength(cloud_metadata = get_metadata_url("cellNexus_lr_signaling_pathway_strength_DEMO.parquet"))
```

### Extract only a subset of genes

This is helpful if just few genes are of interest (e.g ENSG00000134644 (PUM1)), as they can be compared across samples. cellNexus uses ENSEMBL gene ID(s).

```{r get-sce-genes}
single_cell_cpm <-
  metadata |>
  dplyr::filter(
    self_reported_ethnicity == "African" &
      assay == "10x 3' v3" &
      tissue == "lung parenchyma" &
      cell_type_unified_ensemble == "cd14 mono"
  ) |>
  get_single_cell_experiment(assays = "cpm", features = "ENSG00000134644")

single_cell_cpm
```

### Extract the counts as a Seurat object

This convert the H5 SingleCellExperiment to Seurat so it might take long time and occupy a lot of memory depending on how many cells you are requesting.

```{r get-seurat}
seurat_counts <-
  metadata |>
  dplyr::filter(
    self_reported_ethnicity == "African" &
      assay == "10x 3' v3" &
      tissue == "lung parenchyma" &
      cell_type_unified_ensemble == "cd14 mono"
  ) |>
  head() |>
  get_seurat()

seurat_counts
```

By default, data is downloaded to `get_default_cache_dir()` output. If memory is a concern, users can specify a custom cache directory to metadata and counts functions:

## Load metadata from the custom cache directory

```{r custom-cache-metadata, eval=FALSE}
metadata <- get_metadata(cache_directory = "/MY/CUSTOM/PATH")
```


## Query raw counts from the custom cache directory

```{r custom-cache-counts, eval=FALSE}
single_cell_counts <-
  metadata |>
  dplyr::filter(
    self_reported_ethnicity == "African" &
      assay == "10x 3' v3" &
      tissue == "lung parenchyma" &
      cell_type_unified_ensemble == "cd14 mono"
  ) |>
  get_single_cell_experiment(cache_directory = "/MY/CUSTOM/PATH")

single_cell_counts
```

Same strategy can be applied for functions `get_pseuodbulk()` and `get_seurat()` by passing your custom directory character to "cache_directory" parameter.

## Save your `SingleCellExperiment`

The returned `SingleCellExperiment` can be saved with three modalities, as `.rds` or as `HDF5` or as `H5AD`.

### Saving as RDS (fast saving, slow reading)

Saving as `.rds` has the advantage of being fast, and the `.rds` file occupies very little disk space as it only stores the links to the files in your cache.

However it has the disadvantage that for big `SingleCellExperiment` objects, which merge a lot of HDF5 from your `get_single_cell_experiment`, the display and manipulation is going to be slow.
In addition, an `.rds` saved in this way is not portable: you will not be able
to share it with other users.

```{r save-rds, eval=FALSE}
single_cell_counts |>
  saveRDS("single_cell_counts.rds")
```

### Saving as HDF5 (slow saving, fast reading)

Saving as `.hdf5` executes any computation on the `SingleCellExperiment` and writes it to disk as a monolithic `HDF5`.
Once this is done, operations on the `SingleCellExperiment` will be comparatively very fast.
The resulting `.hdf5` file will also be totally portable and sharable.

However this `.hdf5` has the disadvantage of being larger than the corresponding `.rds` as it includes a copy of the count information, and the saving process is going to be slow for large objects.

```{r save-hdf5, eval=FALSE}
# ! IMPORTANT if you save 200K+ cells
HDF5Array::setAutoBlockSize(size = 1e+09)

single_cell_counts |>
  HDF5Array::saveHDF5SummarizedExperiment(
    "single_cell_counts",
    replace = TRUE,
    as.sparse = TRUE,
    verbose = TRUE
  )
```

### Saving as H5AD (slow saving, fast reading)

Saving as `.h5ad` executes any computation on the `SingleCellExperiment` and writes it to disk as a monolithic `H5AD`.
The `H5AD` format is the HDF5 disk representation of the AnnData object and is well-supported in Python.

However this `.h5ad` saving strategy has a bottleneck of handling columns with only NA values of a `SingleCellExperiment` metadata. 

```{r save-h5ad, eval=FALSE}
# ! IMPORTANT if you save 200K+ cells
HDF5Array::setAutoBlockSize(size = 1e+09)

single_cell_counts |>
  anndataR::write_h5ad("single_cell_counts.h5ad",
    compression = "gzip",
    verbose = TRUE
  )
```

## Visualise gene transcription

We can gather all CD14 monocytes cells and plot the distribution of ENSG00000085265 (FCN1) across all tissues

```{r plot-fcn1-disease, eval=FALSE,  results='hide'}
# Plots with styling
counts <- metadata |>
  # Filter and subset
  dplyr::filter(cell_type_unified_ensemble == "cd14 mono") |>
  # Get counts per million for FCN1 gene
  get_single_cell_experiment(assays = "cpm", features = "ENSG00000085265") |>
  suppressMessages() |>
  # Add feature to table
  tidySingleCellExperiment::join_features("ENSG00000085265", shape = "wide") |>
  # Rank x axis
  tibble::as_tibble() |>
  # Rename to gene symbol
  dplyr::rename(FCN1 = ENSG00000085265)

# Plot by disease
counts |>
  dplyr::with_groups(disease, ~ .x |>
    dplyr::mutate(median_count = median(`FCN1`, rm.na = TRUE))) |>
  # Plot
  ggplot(aes(forcats::fct_reorder(disease, median_count, .desc = TRUE), `FCN1`, color = dataset_id)) +
  geom_jitter(shape = ".") +

  # Style
  guides(color = "none") +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) +
  xlab("Disease") +
  ggtitle("FCN1 in CD14 monocytes by disease. Coloured by datasets")
```

```{r fig-fcn1-disease, echo=FALSE}
find_figure("FCN1_disease_plot.png") |>
  knitr::include_graphics()
```

```{r plot-fcn1-tissue, eval=FALSE, results='hide'}
# Plot by tissue
counts |>
  dplyr::with_groups(tissue, ~ .x |>
    dplyr::mutate(median_count = median(`FCN1`, rm.na = TRUE))) |>
  # Plot
  ggplot(aes(
    forcats::fct_reorder(tissue,
      median_count,
      .desc = TRUE
    ),
    `FCN1`,
    color = dataset_id
  )) +
  geom_jitter(shape = ".") +

  # Style
  guides(color = "none") +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) +
  xlab("Tissue") +
  ggtitle("FCN1 in CD14 monocytes by tissue. Colored by datasets") +
  theme(legend.position = "none", axis.text.x = element_text(size = 6.5))
```

```{r fig-fcn1-tissue, echo=FALSE}
find_figure("FCN1_tissue_plot.png") |>
  knitr::include_graphics()
```


## Integrate cloud and local metadata
`cellNexus` not only enables users to query our metadata but also allows integration with your local metadata. Additionally, users can integrate with your metadata stored in the cloud. 

To enable this feature, users must include `file_id_cellNexus_single_cell`
and `atlas_id` (e.g cellxgene/dd-mm-yy) columns in the metadata. See metadata structure in cellNexus::pbmc3k_sce

```{r integrate-local-setup}
# Set up local cache and paths
local_cache <- tempdir()
layer <- "counts"
meta_path <- file.path(local_cache, "pbmc3k_metadata.parquet")
data(pbmc3k_sce)

# Extract and prepare metadata
pbmc3k_metadata <- pbmc3k_sce |>
  S4Vectors::metadata() |>
  purrr::pluck("data") |>
  dplyr::mutate(
    counts_directory = file.path(tempdir(), atlas_id, layer),
    sce_path = file.path(counts_directory, file_id_cellNexus_single_cell)
  )

# Get unique paths
counts_directory <- pbmc3k_metadata |>
  dplyr::pull(counts_directory) |>
  unique()

sce_path <- pbmc3k_metadata |>
  dplyr::pull(sce_path) |>
  unique()

# Create directory structure
dir.create(counts_directory, recursive = TRUE, showWarnings = FALSE)

# Save data to disk
pbmc3k_sce |>
  S4Vectors::metadata() |>
  purrr::pluck("data") |>
  arrow::write_parquet(meta_path)

# Save SCE object
pbmc3k_sce |>
  anndataR::write_h5ad(sce_path, compression = "gzip", mode = "w")
```

```{r integrate-local-query}
# A cellNexus file
file_id_from_cloud <- "e52795dec7b626b6276b867d55328d9f___1.h5ad"
file_id_local <- basename(sce_path)

get_metadata(
  cloud_metadata = METADATA_URL,
  local_metadata = meta_path,
  cache_directory = local_cache
) |>
  # For illustration purpose, only filter a selected cloud metadata and the saved metadata
  dplyr::filter(file_id_cellNexus_single_cell %in% c(file_id_from_cloud, file_id_local)) |>
  dplyr::select(cell_id, sample_id, dataset_id, cell_type_unified_ensemble, atlas_id, file_id_cellNexus_single_cell) |>
  get_single_cell_experiment(cache_directory = local_cache)
```



# Cell metadata

Field definitions for the CELLxGENE schema follow the
[CELLxGENE schema 5.1.0](https://github.com/chanzuckerberg/single-cell-curation/blob/main/schema/5.1.0/schema.md).

## Dataset-level columns

| Column | Description |
|--------|-------------|
| `dataset_id` | Primary dataset identifier in the atlas. |
| `collection_id` | Parent collection grouping related datasets. |
| `citation` | Citation or reference for the dataset. |
| `dataset_version_id` | Versioned identifier for a dataset release. |
| `explorer_url` | URL to browse the dataset in a public explorer. |
| `filesize` | Reported size of source data files. |
| `filetype` | Storage format of source expression files. |
| `published_at` | Publication timestamp for dataset release. |
| `raw_data_location` | Location of raw data upstream. |
| `revised_at` | Last revision timestamp for dataset metadata. |
| `schema_version` | CELLxGENE schema version for this record. |
| `suspension_type` | Suspension type (e.g. cell vs nucleus). |
| `title` | Dataset title. |
| `url` | Dataset or resource URL. |
| `donor_id` | Donor or participant identifier. |

## Sample-level columns

| Column | Description |
|--------|-------------|
| `assay` | Assay or platform label. |
| `assay_ontology_term_id` | Ontology identifier for `assay`. |
| `development_stage` | Developmental stage label. |
| `development_stage_ontology_term_id` | Ontology identifier for `development_stage`. |
| `self_reported_ethnicity` | Self-reported ancestry or ethnicity label. |
| `self_reported_ethnicity_ontology_term_id` | Ontology identifier for self-reported ethnicity. |
| `experiment___` | Upstream experiment grouping variable. |
| `organism` | Organism name (e.g. human). |
| `organism_ontology_term_id` | Ontology identifier for `organism`. |
| `sex` | Recorded biological sex label. |
| `sex_ontology_term_id` | Ontology identifier for `sex`. |
| `tissue` | Tissue label. |
| `tissue_type` | Tissue class (e.g. tissue vs organoid). |
| `tissue_ontology_term_id` | Ontology identifier for `tissue`. |
| `disease` | Disease or condition annotation. |
| `disease_ontology_term_id` | Ontology identifier for `disease`. |
| `is_primary_data` | Whether observations are marked as primary upstream. |

## Cell-level columns

| Column | Description |
|--------|-------------|
| `cell_id` | Unique cell barcode or identifier. |
| `cell_type` | Cell-type label from source or curation. |
| `cell_type_ontology_term_id` | Ontology identifier for `cell_type`. |
| `cell_annotation_azimuth_l2` | Azimuth level-2 annotation used in harmonisation. |
| `cell_annotation_blueprint_singler` | SingleR annotation using the Blueprint reference. |
| `observation_joinid` | Join key linking metadata and expression records. |
| `empty_droplet` | Quality-control flag for probable empty droplets. |
| `alive` | Quality-control flag for viable cells (e.g. mitochondrial signal). |
| `scDblFinder.class` | Doublet classification from `scDblFinder`. |
| `is_immune` | Curated flag for immune-cell context. |

## Harmonised and custom columns (cellNexus)

Through harmonisation and curation, cellNexus adds columns that are not present in the original CELLxGENE metadata alone.

| Column | Description |
|--------|-------------|
| `cell_count` | Number of cells in a dataset. |
| `feature_count` | Number of genes in a dataset. |
| `age_days` | Donor age in days. |
| `tissue_groups` | Coarse tissue grouping for analysis. |
| `empty_droplet` | Whether a cell is called an empty droplet from expressed-gene count per sample (default threshold 200; targeted panels may differ). |
| `alive` | Whether a cell passes viability / mitochondrial QC. |
| `scDblFinder.class` | Doublet, singlet, or unknown (`scDblFinder` default parameters). |
| `cell_type_unified_ensemble` | Consensus immune identity from Azimuth and SingleR (Blueprint, Monaco). |
| `cell_annotation_azimuth_l2` | Azimuth cell annotation. |
| `cell_annotation_blueprint_singler` | SingleR annotation (Blueprint). |
| `cell_annotation_blueprint_monaco` | SingleR annotation (Monaco). |
| `sample_heuristic` | Internal sample subdivision helper. |
| `file_id_cellNexus_single_cell` | Internal file id for single-cell layers. |
| `file_id_cellNexus_pseudobulk` | Internal file id for pseudobulk layers. |
| `sample_id` | Harmonised sample identifier. |
| `nCount_RNA` | Total RNA counts per cell (sample-aware). |
| `nFeature_expressed_in_sample` | Number of expressed features per cell. |
| `ethnicity_flagging_score` | Supporting score for ethnicity imputation. |
| `low_confidence_ethnicity` | Supporting flag for low-confidence ethnicity calls. |
| `.aggregated_cells` | Post-QC cells combined into each pseudobulk sample. |
| `imputed_ethnicity` | Imputed ethnicity label. |
| `atlas_id` | cellNexus atlas release identifier (internal use). |

# RNA abundance

The `counts` assay includes RNA abundance in the positive real scale (not transformed with non-linear functions, e.g. log sqrt). Originally CELLxGENE include a mix of scales and transformations specified in the `x_normalization` column.

The `cpm` assay includes counts per million.

The `sct` assay includes normalised counts by `sctranform`.

# Other representations

The `rank` assay is the representation of each cell’s gene expression profile where genes are ranked by expression intensity.

The `pseudobulk` assay includes aggregated RNA abundance for sample and cell type combination.


# Session Info
```{r session-info}
sessionInfo()
```
